Predicting Ensemble Location and Width in Binaural Recordings of Music with Convolutional Neural Networks

### 2024-06-23 paper/main.tex
Update reference styles, table format and software details in the paper

\usepackage[english]{babel}
\usepackage{authblk}
\usepackage[a4paper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{gensymb}
\usepackage{textcomp}
\usepackage[backend=biber, style=nature]{biblatex}
\usepackage{csquotes}
\usepackage[toc,page]{appendix}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{multirow}
\usepackage{tabularx}

@@

The human auditory system demonstrates exceptional proficiency in segregating, localizing, and interpreting diverse auditory signals, despite being limited to two ears. This capability arises from the intricate analysis of temporal, amplitude, and spectral disparities, a process termed binaural hearing \cite{blauert_spatial_1996}, which enables precise localization of sound sources in complex auditory environments. An important advantage of binaural hearing is demonstrated by the `cocktail party effect', which showcases the system's capacity to concentrate on individual sounds while suppressing background noise \cite{cherry_experiments_1953}. An understanding of the auditory system is essential for comprehending its limits and creating more immersive binaural experiences for entertainment purposes \cite{zhang_surround_2017}. It also helps to improve hearing aid devices by enhancing auditory signal reception and improving spatial awareness \cite{hirsh_binaural_1950, thiemann_speech_2016}.

The advancement of sophisticated machine learning techniques, particularly deep learning networks, has spurred an intriguing exploration into the extent to which these tools can emulate the human auditory system. This exploration proceeds without reliance on the advanced spatial audio feature engineering traditionally used in audio source localization, as evidenced by references \cite{yang_deepear_2022, vera-diaz_towards_2018, pang_multitask_2019}. While the application of convolutional neural networks (CNNs) \cite{lecun_handwritten_1989} to audio signals --- especially in conjunction with spectrograms \cite{thomas_analyzing_2014, espi_exploiting_2015} or other techniques \cite{abdel-hamid_applying_2012, sainath_deep_2013} --- is not novel, it has been successfully employed in prior studies. Building on these foundations, this study developed an audio localization method by creating a multi-task CNN model.

Inspired by the fact, that humans tend to localize sound sources in groups rather than individually \cite{bregman_auditory_1990, rumsey_spatial_2002}, the objective of the proposed model is to predict ensemble location and width instead of positions of individual sources. This study is unique as it not only conceptualized the method but also tested it on a vast, real-life music corpus of 23,040 binaural excerpts that were synthesized using 192 multi-track music recordings and 30 publicly available head-related transfer functions (HRTFs) from various sources. The music recordings encompassed various genres, including rock, jazz, pop, and classical music.

The findings demonstrate that this method is effective in accurately predicting the spatial characteristics of sound sources in near-real-world scenarios. Furthermore, this paper presents an experiment framework that allows for the objective measurement of a binaural localization technique in an objective manner, utilizing a large-scale dataset synthesized from real-world examples of music signals (for other use cases for this framework, see \cite{antoniuk2023blind, zielinski_automatic_2022, zielinski_spatial_2022, zielinski_comparison_2020}). One of the key advantages of the proposed method is that it does not assume the number of audio sources. However, a significant limitation of this study is the lack of reverberation in the synthesized recordings, which represents a material for future research.

@@

The majority of existing literature on the subject of sound source localization employs techniques that leverage the advantages of microphone arrays with more than two channels \cite{kaveh_statistical_1986, pavlidi_real-time_2012, pan_multi-tone_2021, hahmann_sound_2022, chung_sound_2022, liu_sound_2022}. In the context of sound source localization in binaural signals, the predominant focus of research is on the identification of individual sound sources, rather than groups of sounds \cite{dietz_auditory_2011, may_probabilistic_2011, may_binaural_2012, woodruff_binaural_2012, may_robust_2015, ma16c_interspeech, ma_exploiting_2017, benaroya_binaural_2018}. In terms of source direction of arrival (DOA) methods, the majority of research assumes a fixed number of sound sources \cite{pang_multitask_2019, vecchiotti19, ma_exploiting_2017, woodruff_binaural_2012, s_spatiogram_2021}, which limits its practical applications as this information is rarely known in real-life binaural recordings. Moreover, the majority of studies have focused on relatively homogeneous signals, namely speech \cite{dietz_auditory_2011, may_probabilistic_2011, may_binaural_2012, woodruff_binaural_2012, may_robust_2015, ma16c_interspeech, ma_exploiting_2017, benaroya_binaural_2018, wang_binaural_2020, liu_multiple_2018, yang_deepear_2022, ma_robust_2018}.

In contrast to the aforementioned studies, the proposed method is not constrained by the number of sources. Furthermore, the approach is not limited to speech and has been applied to a wide range of musical datasets, including instruments and vocals. In contrast to previous studies that primarily focused on individual sources, the proposed method does not aim to separate them, but rather considers them as a group, or in this case, a musical ensemble. This approach is similar to how the real musical ensembles are arranged on stage, thereby emphasizing practical applications in live settings. To the authors' knowledge, this is one of the first methods to localize ensemble width (see \cite{antoniuk2023blind} for the previous ensemble-width-related study), and the first to localize both ensemble position and width simultaneously using a multi-task model.

Sound localization methods can be classified into two categories based on the implementation of their underlying algorithms: glass-box (e.g., \cite{dietz_auditory_2011, may_probabilistic_2011, may_binaural_2012, may_robust_2015, woodruff_binaural_2012, ma16c_interspeech, ma_exploiting_2017, ma_robust_2018}) and black-box (e.g., \cite{vecchiotti19, vera-diaz_towards_2018, yang_deepear_2022}). Glass-box methods, more traditional in the literature, rely on manually designed algorithms that mimic the auditory system to explicitly extract key features for location prediction, such as interaural level differences, interaural time differences, interaural coherence, or interaural phase differences (see \cite{blauert_spatial_1996} for feature descriptions). An example of one of the most advanced auditory models that is able to extract such features was developed by the Two!Ears project \cite{raake_computational_nodate}.

In contrast, black-box methods generally employ minimal feature engineering, relying instead on advanced machine learning techniques to autonomously extract features and make predictions. These methods, while effective, do not explicitly reveal the features they extract nor necessarily mimic the human auditory system. This opacity and the unpredictable nature of the outcomes, coupled with their dependency on deep neural networks that have extensive learning parameters, require the use of very large datasets for development and evaluation. Such datasets often include thousands of examples, as seen in the TIMIT corpus with 6300 examples \cite{garofolo_john_s_timit_1993} used in multiple studies \cite{yang_deepear_2022, benaroya_binaural_2018, wang_binaural_2020, vecchiotti19, ma_exploiting_2017, pang_multitask_2019, ma_robust_2018, may_robust_2015}, or custom corpora comprising hundreds of thousands of recordings \cite{antoniuk2023blind, zielinski_automatic_2022, zielinski_spatial_2022, zielinski_comparison_2020}. This poses a significant challenge in assembling a sufficiently large and diverse collection of labeled binaural recordings. However, this challenge can be addressed through the synthesis of binaural sounds, as demonstrated in various studies \cite{antoniuk2023blind, zielinski_automatic_2022, zielinski_spatial_2022, zielinski_comparison_2020,yang_deepear_2022, ma_robust_2018} and further elaborated upon in Section \ref{subsec:synthesis} of this paper.

@@

The objective of the model developed in this study is to predict the ensemble location ($\theta$) and width ($\omega$), as illustrated in Figure \ref{fig:scene}. An ensemble is defined as a group of audio point sources positioned on a circle around the listener on a virtual acoustic scene with equal distance to the listener. The location of source $i$ is denoted by $\theta_i$. The ensemble width ($\omega$) is defined as the angular width between two extreme point sources ($max_i(\theta_i)-min_i(\theta_i)$), while the ensemble location, designated by $\theta$, represents the middle angle between two extreme sound sources ($(max_i(\theta_i)+min_i(\theta_i))/2$). For the purposes of this study, the locations of the sources were limited to the frontal hemisphere only, i.e. $\theta\in[-45\degree, 45\degree]$, $\omega\in[0\degree, 90\degree]$, as this range encompasses the majority of real-world recording scenarios. It should be noted that although humans possess some limited abilities to localize sound sources in the vertical plane, in this study all sources are placed in the horizontal plane, at the height of the listener. This covers the majority of cases for real-world recordings (see \cite{ma_robust_2018, zielinski_spatial_2022} for similar studies that cover top-down discrimination).

@@

The experiments conducted in this study involved 23,040 binaural recordings of music. The binaural recordings were synthesized using 192 multi-track publicly-available music recordings \cite{noauthor_mixing_nodate} and 30 HRTF databases (see Table \ref{table:hrtfs} in \nameref{appendix:a_hrtf} for a detailed list of HRTF databases used in this study). The number of tracks in multi-track recording ranged from 5 to 62, with median of 9. For each pair of a multi-track recording and an HRTF database, four binaural recordings were synthesized with different random ensemble parameters, namely location $\theta$ and width $\omega$, as defined in Section \ref{subsec:ensemble_definition}. Both parameters were drawn from a uniform random distribution. Furthermore, the tracks of the input multi-track recordings were randomly assigned to sound source positions ($\theta_i$) to enhance the diversity of the final binaural corpora. Before the synthesis, the tracks were equalized to $-23$ LUFS, in accordance with the ITU-R BS.1770 recommendation \cite{noauthor_recommendation_nodate}.

The binaural recordings were obtained in this study using the binaural synthesis procedure, known as binauralization, whose aim was to simulate the positions of sound sources within a virtual acoustic environment \cite{blauert_spatial_1996}. This was achieved by convolving multi-track signals with head-related impulse responses from a specified head-related transfer function (HRTF) database. The resulting binaural output signal $y_c[n]$ for each stereo channel $c$ (left or right) at sample $n$ is given by the following equation:
\begin{equation}
  y_c[n] = \sum_{i=1}^{N} \sum_{k=0}^{K-1} x_i[k] \times h_{c,\theta_i}[n-k] ,
\end{equation}
where $x_i$ represents the signal of an individual sound source $i$ from the input music recording and $h_{c,\theta_i}$ denotes the head-related impulse response for channel $c$ at location $\theta_i$ of source track $i$.

@@

% The procedure proposed in this study did not

% Zamiast tego dac obrazek z glowami i spektrogramami, lewy i prawi i roznica, sprawdzic czy to ma sens, potencjalnie usunac
% \begin{figure}[ht]
%   \centering
%   \includegraphics[width=\linewidth]{../figures/samples.pdf}
%   \caption{\label{fig:samples}Example magnitude spectrograms from randomly selected synthesized binaural music recordings, used as input data for training and evaluating the convolutional neural network (CNN) in this study. }
% \end{figure}

% Figure \ref{fig:samples} displays example spectrograms from four distinct binaural recordings, exhibiting a broad bandwidth and the diversity of the signals used in this study

Prior to input into the model, the binaural recordings of music were transformed into magnitude spectrograms. It is important to note that the original spectrograms were preserved in a floating-point precision matrix format, which prevented any loss of information due to precision conversion. The spectrograms were calculated using the Fast Fourier Transform (FFT) algorithm, with a limitation of 150 bands, spaced linearly from 100 Hz to 16 kHz. Additionally, a Hamming window of 40 ms was applied to each frame of the signal, resulting in a total of 349 time frames. This procedure was conducted for both the left and right channels, yielding two spectrograms for each binaural sample. Each sample was represented by a matrix of dimensions $2 \times 349 \times 150$. This method parallels the procedure presented in \cite{zielinski_automatic_2022}.

@@

The network topology employed in this study, depicted in Figure \ref{fig:architecture}, was significantly influenced by the AlexNet convolutional neural network \cite{krizhevsky_imagenet_2012}. Although AlexNet was originally designed for image classification, this study adapted it for the audio prediction task by converting binaural recordings into magnitude spectrograms, as described in Section \ref{subsec:feature_extraction}. This conversion allowed the spectrograms to be treated as images, enabling their use in a standard image-prediction-like task.

@@

Despite the availability of widely used techniques for addressing overfitting, such as the Dropout Layer \cite{srivastava_dropout_nodate}, and for accelerating training, such as Batch Normalization \cite{ioffe_batch_2015}, neither technique was employed in this study due to their observed ineffectiveness for the specific prediction task at hand. Instead, a Global Average Pooling Layer \cite{lin_network_2014} was utilized, known for its proven capabilities in reducing overfitting. In this particular task, the inclusion of this layer significantly reduced overfitting, lowering the final mean absolute error score by $0.83\degree$ (average across 10 trials) compared to configurations using a Flattening Layer.

@@

Monte-Carlo cross validation (without replacement).

@@

{\renewcommand{\arraystretch}{1.1}
\begin{table}[!h]
  \begin{center}
    \caption{\label{table:hrtfs}List of HRTF sets used to synthesize binaural audio excerpts}
    \begin{tabularx}{\linewidth}{
      |>{\hsize=0.3\hsize}X|
      >{\hsize=0.7\hsize}X|
      >{\raggedright\hsize=1.3\hsize}X|
      >{\raggedright\hsize=0.6\hsize}X|
      >{\raggedright\arraybackslash\hsize=2.1\hsize}X|
      >{\hsize=1\hsize}X|

@@

      \cline{1-6}

@@

      \cline{1-6}
      1.           & Human         & Human subject                             & 1.2                     & \multirow{2}{\hsize}{RWTH Aachen University \cite{braren_high-resolution_2020}}                                                                             & \multirow{2}{\hsize}{AACHEN}    \\
      \cline{1-4}

@@

      \cline{1-6}
      3.           & Human         & Subject 2                                 & 1.2                     & \multirow{4}{\hsize}{Austrian Academy of Sciences \cite{noauthor_hrtf-database_nodate}}                                                                     & \multirow{4}{\hsize}{ARI}       \\
      \cline{1-4}

@@

      \cline{1-4}

@@

      \cline{1-6}
      7.           & Human         & Subject 012                               & 1                       & \multirow{3}{\hsize}{CIPIC Interface Laboratory, University of California \cite{algazi_cipic_2001}}                                                         & \multirow{3}{\hsize}{CIPIC}     \\
      \cline{1-4}

@@

      \cline{1-6}
      10.          & Artificial    & Neumann KU 100                            & 0.9                     & NASA (2007) \cite{andreopoulou_inter-laboratory_2015}                                                                                                       & \multirow{2}{\hsize}{CLUBFRITZ} \\
      \cline{1-6}
      11.          & Artificial    & Neumann KU 100                            & 1.5                     & Helsinki University of Technology (2009) \cite{andreopoulou_inter-laboratory_2015}                                                                          &                                 \\
      \cline{1-6}
      12.          & Artificial    & FABIAN                                    & 1.47                    & \multirow{4}{\hsize}{Technical University Berlin, Huawei Technologies, Munich Research Centre, Sennheiser Electronic \cite{brinkmann_cross-evaluated_2019}} & \multirow{4}{\hsize}{HUTUBS}    \\
      \cline{1-4}

@@

                   &               &                                           &                         &                                                                                                                                                             &                                 \\
      \cline{1-6}
      15.          & Human         & Subject 1003                              & 1.95                    & \multirow{2}{\hsize}{IRCAM, AKG \cite{noauthor_listen_2023}}                                                                                                & \multirow{2}{\hsize}{LISTEN}    \\
      \cline{1-4}

@@

                   &               &                                           &                         &                                                                                                                                                             &                                 \\
      \cline{1-6}
    \end{tabularx}

>>>>

\usepackage[english]{babel}
\usepackage{authblk}
\usepackage[a4paper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{gensymb}
\usepackage{textcomp}
\usepackage[backend=biber, style=apa, maxcitenames=1, uniquelist=false, url=false]{biblatex}
\usepackage{csquotes}
\usepackage[toc,page]{appendix}
\usepackage[colorlinks=true, allcolors=blue, breaklinks=true]{hyperref}
\usepackage{xurl}
\usepackage{multirow}
\usepackage{tabularx}
\usepackage{ltablex}
\usepackage{color}
\usepackage{tabularray}

\defbibenvironment{bibliography}
  {\begin{enumerate}}
  {\end{enumerate}}
  {\item}
\AtEveryBibitem{%
  \clearfield{note}%
}

% \AtEveryCitekey{\iffootnote{\csxdef{blx@footnotemark}{}\csxdef{blx@shortfootnotemark}{}}{}}

% \DeclareCiteCommand{\cite}
%   {\usebibmacro{prenote}}
%   {\usebibmacro{citeindex}%
%    \printtext[bibhyperref]{\usebibmacro{cite}}
%   }
%   {\multicitedelim}
%   {\usebibmacro{postnote}}

%   \DeclareCiteCommand{\parencite}
%   {\usebibmacro{prenote}}
%   {\usebibmacro{citeindex}%
%    \printtext[bibhyperref]{\bibopenparen\usebibmacro{cite}\bibcloseparen}
%   }
%   {\multicitedelim}
%   {\usebibmacro{postnote}}

% % Redefine \parencites to handle multiple citations in one set of parentheses
% \DeclareCiteCommand{\parencites}
%   {\usebibmacro{prenote}}
%   {\usebibmacro{citeindex}%
%    \printtext[bibhyperref]{\bibopenparen\usebibmacro{cite}\bibcloseparen}
%   }
%   {\multicitedelim}
%   {\usebibmacro{postnote}}

@@

The human auditory system demonstrates exceptional proficiency in segregating, localizing, and interpreting diverse auditory signals, despite being limited to two ears. This capability arises from the intricate analysis of temporal, amplitude, and spectral disparities, a process termed binaural hearing \parencite{blauert_spatial_1996}, which enables precise localization of sound sources in complex auditory environments. An important advantage of binaural hearing is demonstrated by the `cocktail party effect', which showcases the system's capacity to concentrate on individual sounds while suppressing background noise \parencite{cherry_experiments_1953}. An understanding of the auditory system is essential for comprehending its limits and creating more immersive binaural experiences for entertainment purposes \parencite{zhang_surround_2017}. It also helps to improve hearing aid devices by enhancing auditory signal reception and improving spatial awareness \parencite{hirsh_binaural_1950, thiemann_speech_2016}.

The advancement of sophisticated machine learning techniques, particularly deep learning networks, has prompted an intriguing exploration into the extent to which these tools can emulate the human auditory system. This exploration proceeds without reliance on the advanced spatial audio feature engineering traditionally used in audio source localization, as evidenced by some prominent recent examples \parencite{yang_deepear_2022, vera-diaz_towards_2018, pang_multitask_2019}. While the application of convolutional neural networks (CNNs) \parencite{lecun_handwritten_1989} --- traditionally employed in the visual domain --- to audio signals, particularly in conjunction with spectrograms \parencite{thomas_analyzing_2014, espi_exploiting_2015} or other feature engineering techniques \parencite{abdel-hamid_applying_2012, sainath_deep_2013}, is not novel. However, these approaches are continually being refined and adapted to the audio realm. Building on these foundations, this study developed an audio localization method by constructing a spectrogram-based multi-task CNN model.

Inspired by the fact, that humans tend to localize sound sources in groups rather than individually \parencite{bregman_auditory_1990, rumsey_spatial_2002}, the objective of the proposed model is to predict ensemble location and width instead of positions of individual sources. This study is unique as it not only conceptualized the method but also tested it on a vast, real-life music corpus of 23,040 binaural excerpts that were synthesized using 192 multi-track music recordings \parencite{senior_mixing_2023} and 30 publicly available head-related transfer functions (HRTFs) from various sources (see Table \ref{table:hrtfs} in \ref{appendix:a_hrtf}). The music recordings encompassed various genres, including rock, jazz, pop, and classical music.

The findings demonstrate that this method is effective in accurately predicting the spatial characteristics of sound sources in near-real-world scenarios. Furthermore, this paper demonstrates an experimental framework that facilitates the objective measurement of a binaural localization technique, employing a large-scale dataset synthesized from real-world music signal (for applications of similar frameworks, see: \cite{antoniuk_blind_2023, zielinski_automatic_2022, zielinski_spatial_2022, zielinski_comparison_2020}). One of the key advantages of the proposed method is that it does not assume the number of audio sources. However, significant limitations of this study include the absence of reverberation in the synthesized recordings and the method's inapplicability to real-time scenarios --- both are critical areas for future research.

@@

The majority of existing literature on the subject of sound source localization employs techniques that leverage the advantages of microphone arrays with more than two channels \parencite{kaveh_statistical_1986, pavlidi_real-time_2012, pan_multi-tone_2021, hahmann_sound_2022, chung_sound_2022, liu_sound_2022}. In the context of sound source localization in binaural signals, the predominant focus of research is on the identification of individual sound sources, rather than groups of sounds \parencite{dietz_auditory_2011, may_probabilistic_2011, may_binaural_2012, woodruff_binaural_2012, may_robust_2015, ma16c_interspeech, ma_exploiting_2017, benaroya_binaural_2018}. In terms of source direction of arrival (DOA) methods, the majority of research assumes a fixed number of sound sources \parencite{pang_multitask_2019, vera-diaz_towards_2018, ma_exploiting_2017, woodruff_binaural_2012, arthi_spatiogram_2021}, which limits its practical applications as this information is rarely known in real-life binaural recordings. Moreover, the majority of studies have focused on relatively homogeneous signals, namely speech \parencite{dietz_auditory_2011, may_probabilistic_2011, may_binaural_2012, woodruff_binaural_2012, may_robust_2015, ma16c_interspeech, ma_exploiting_2017, benaroya_binaural_2018, wang_binaural_2020, liu_multiple_2018, yang_deepear_2022, ma_robust_2018}.

In contrast to the aforementioned studies, the proposed method is not constrained by the number of sources. Furthermore, the approach is not limited to speech and has been applied to a wide range of musical datasets, including instruments and vocals. In contrast to previous studies that primarily focused on individual sources, the proposed method does not aim to separate them, but rather considers them as a group, or in this case, a musical ensemble. This approach is similar to how the real musical ensembles are arranged on stage, thereby emphasizing practical applications in live settings. To the authors' knowledge, this is one of the first methods to localize ensemble width (see \cite{antoniuk_blind_2023} for the previous ensemble-width-related study), and the first to localize both ensemble position and width simultaneously using a multi-task model.

Sound localization methods can be classified into two categories based on the implementation of their underlying algorithms: glass-box (e.g., \parencite{dietz_auditory_2011, may_probabilistic_2011, may_binaural_2012, may_robust_2015, woodruff_binaural_2012, ma16c_interspeech, ma_exploiting_2017, ma_robust_2018}) and black-box (e.g., \parencite{vera-diaz_towards_2018, vera-diaz_towards_2018, yang_deepear_2022}). Glass-box methods, more traditional in the literature, rely on manually designed algorithms that mimic the auditory system to explicitly extract key features for location prediction, such as interaural level differences, interaural time differences, interaural coherence, or interaural phase differences (see \cite{blauert_spatial_1996} for feature descriptions). An example of one of the most advanced auditory models that is able to extract such features was developed by the Two!Ears project \parencite{raake_computational_nodate}.

In contrast, black-box methods generally employ minimal feature engineering, relying instead on advanced machine learning techniques to autonomously extract features and make predictions. These methods, while effective, do not explicitly reveal the features they extract nor necessarily mimic the human auditory system. This opacity and the unpredictable nature of the outcomes, coupled with their dependency on deep neural networks that have extensive learning parameters, require the use of very large datasets for development and evaluation. Such datasets often include thousands of examples, as seen in the TIMIT corpus with 6300 examples \parencite{garofolo_darpa_1993} used in multiple studies \parencite{yang_deepear_2022, benaroya_binaural_2018, wang_binaural_2020, vera-diaz_towards_2018, ma_exploiting_2017, pang_multitask_2019, ma_robust_2018, may_robust_2015}, or custom corpora comprising hundreds of thousands of recordings \parencite{antoniuk_blind_2023, zielinski_automatic_2022, zielinski_spatial_2022, zielinski_comparison_2020}. This poses a significant challenge in assembling a sufficiently large and diverse collection of labeled binaural recordings. However, this challenge can be addressed through the synthesis of binaural sounds, as demonstrated in various studies \parencite{antoniuk_blind_2023, zielinski_automatic_2022, zielinski_spatial_2022, zielinski_comparison_2020,yang_deepear_2022, ma_robust_2018} and further elaborated upon in Section \ref{subsec:synthesis} of this paper.

@@

The objective of the model developed in this study is to predict the ensemble location ($\theta$) and width ($\omega$), as illustrated in Figure \ref{fig:scene}. An ensemble is defined as a group of audio point sources positioned on a circle around the listener on a virtual acoustic scene with equal distance to the listener. The location of source $i$ is denoted by $\theta_i$. The ensemble width ($\omega$) is defined as the angular width between two extreme point sources ($max_i(\theta_i)-min_i(\theta_i)$), while the ensemble location, designated by $\theta$, represents the middle angle between two extreme sound sources ($(max_i(\theta_i)+min_i(\theta_i))/2$). For the purposes of this study, the locations of the sources were limited to the frontal hemisphere only, i.e. $\theta\in[-45\degree, 45\degree]$, $\omega\in[0\degree, 90\degree]$, as this range encompasses the majority of real-world recording scenarios. It should be noted that although humans possess some limited abilities to localize sound sources in the vertical plane, in this study all sources are placed in the horizontal plane, at the height of the listener. This covers the majority of cases for real-world recordings (see \cite{ma_robust_2018, zielinski_spatial_2022} for related studies that cover top-down discrimination).

@@

The experiments conducted in this study involved 23,040 binaural recordings of music. The binaural recordings were synthesized using 192 multi-track publicly-available music recordings \parencite{senior_mixing_2023} and 30 HRTF databases (see Table \ref{table:hrtfs} in \nameref{appendix:a_hrtf} for a detailed list of HRTF databases used in this study). The number of tracks in multi-track recording ranged from 5 to 62, with median of 9. For each pair of a multi-track recording and an HRTF database, four binaural recordings were synthesized with different random ensemble parameters, namely location $\theta$ and width $\omega$, as defined in Section \ref{subsec:ensemble_definition}. Both parameters were drawn from a uniform random distribution. Furthermore, the tracks of the input multi-track recordings were randomly assigned to sound source positions ($\theta_i$) to enhance the diversity of the final binaural corpora. Before the synthesis, the tracks were equalized to $-23$ LUFS, in accordance with the \parencite{noauthor_itu-r_2023} recommendation.

The binaural recordings were obtained in this study using the binaural synthesis procedure, known as binauralization, whose aim was to simulate the positions of sound sources within a virtual acoustic environment \parencite{blauert_spatial_1996}. This was achieved by convolving multi-track signals with head-related impulse responses from a specified head-related transfer function (HRTF) database. The resulting binaural output signal $y_c[n]$ for each stereo channel $c$ (left or right) at sample $n$ is given by the following equation:
\begin{equation}
  y_c[n] = \sum_{i=1}^{N} \sum_{k=0}^{K-1} x_i[k] \times h_{c,\theta_i}[n-k] ,
\end{equation}
where $x_i$ represents the signal of an individual sound source $i$ from the input music recording and $h_{c,\theta_i}$ denotes the head-related impulse response for channel $c$ at location $\theta_i$ of source track $i$.

@@

% The procedure proposed in this study did not 
% Zamiast tego dac obrazek z glowami i spektrogramami, lewy i prawi i roznica, sprawdzic czy to ma sens, potencjalnie usunac
% \begin{figure}[ht]
%   \centering
%   \includegraphics[width=\linewidth]{../figures/samples.pdf}
%   \caption{\label{fig:samples}Example magnitude spectrograms from randomly selected synthesized binaural music recordings, used as input data for training and evaluating the convolutional neural network (CNN) in this study. }
% \end{figure}
% Figure \ref{fig:samples} displays example spectrograms from four distinct binaural recordings, exhibiting a broad bandwidth and the diversity of the signals used in this study

Prior to input into the model, the binaural recordings of music were transformed into magnitude spectrograms. Although spectrograms do not directly provide information that can be translated into ensemble features, especially ensemble width, the goal of this task was to reduce the number of independent variables compared to the raw audio signal by extracting more compressed and informative data in the frequency domain. This step was necessary to decrease the likelihood of overfitting, reduce the number of examples needed to train the model, and thereby lower the overall computational power requirements. It is worth mentioning, however, that recent methods have shown that CNNs can be suitable for end-to-end audio localization without the spectrogram extraction step \parencite{vera-diaz_towards_2018, vecchiotti_end--end_2019}.

On the computational side, the spectrograms were treated as three-dimensional floating-point precision feature matrices. The spectrograms were calculated using the Fast Fourier Transform (FFT) algorithm, with a limitation of 150 bands spaced linearly from 100 Hz to 16 kHz. Additionally, a Hamming window of 40 ms was applied to each frame of the signal, resulting in a total of 349 time frames. This procedure was conducted for both the left and right channels, yielding two spectrograms for each binaural sample. Each sample was represented by a matrix of dimensions $2 \times 349 \times 150$. This method parallels the procedure presented by \cite{zielinski_automatic_2022}.

@@

The network topology employed in this study, depicted in Figure \ref{fig:architecture}, was significantly influenced by the AlexNet convolutional neural network demonstrated by \cite{krizhevsky_imagenet_2012}. Although AlexNet was originally designed for image classification, this study adapted it for the audio prediction task by converting binaural recordings into magnitude spectrograms, as described in Section \ref{subsec:feature_extraction}. This conversion allowed the spectrograms to be treated as images, enabling their use in a standard image-prediction-like task.

@@

The topology finalized in this study was selected based on the highest prediction quality observed on the validation dataset. Despite the existence of numerous algorithms for automatic topology selection \parencite{branke_evolutionary_1995, zhang_finding_2018, miikkulainen_evolving_2017, stanley_evolving_2002, shafiee_deep_2016}, the final topology selected in this study was determined manually, based on its superior prediction quality observed on the validation dataset. This approach was necessitated primarily by the high computational demands compared to the available resources.

Various architectural configurations were assessed, varying key parameters such as the number of convolutional layers (from 1 to 5), the number of classification units (from 1 to 5), the inclusion or exclusion of Max Pooling Layers after each Convolution Layer, the number of filters within the Convolution Layers, the dimensions of these filters ($2 \times 2$, $2 \times 3$, or $3 \times 3$), the stride size, and the dimensions of the Max Pooling Layers ($2 \times 2$, $2 \times 3$, or $3 \times 3$). Although a detailed account of the selection process for the topology is beyond the primary scope of this study --- particularly considering the task's complexity --- it was concluded that the Convolutional Neural Network (CNN) model is robust against variations in the assessed topologies. The differences in mean prediction error among the configurations were minimal, typically less than $1^\circ$ for most configurations. Among the many tested topologies that yielded almost identical errors, the simplest one was chosen to optimize both performance efficiency and model simplicity.

Despite the availability of widely used techniques for addressing overfitting, such as the Dropout Layer \parencite{srivastava_dropout_2014}, and for accelerating training, such as Batch Normalization \parencite{ioffe_batch_2015}, neither technique was employed in this study due to their observed ineffectiveness for the specific prediction task at hand. Instead, a Global Average Pooling Layer \parencite{lin_network_2013} was utilized, known for its proven capabilities in reducing overfitting. In this particular task, the inclusion of this layer significantly reduced overfitting, lowering the final mean absolute error score by $0.83\degree$ (average across 10 trials) compared to configurations using a Flattening Layer.

@@

The topology described previously resulted in a model with 216,562 learning parameters. This model was trained 10 times from scratch, employing the Monte Carlo cross-validation method \parencite{kuhn_applied_2013}. For each iteration, the entire dataset was randomly divided into two parts: a development set containing two-thirds of the dataset (15,360 recordings) for model construction, and a test set comprising the remaining one-third (7,680 recordings) for evaluation. The development set was further subdivided into training and validation subsets at a 7:1 ratio, with 13,440 recordings allocated to the training subset and 1,920 recordings to the validation subset.

The training subset was employed to update the model's learning parameters, while the validation subset was utilized for implementing the early stopping technique \parencite{pocock_practical_1989, morgan_generalization_1989} in conjunction with model checkpointing \parencite{eisenman_check-n-run_2020}. These methods played a crucial role in selecting the model that demonstrated the best generalization capabilities. The test subset, containing data not previously encountered in the training or validation phases, was used exclusively for performance assessment once per iteration. This process was repeated to collect 10 mean absolute errors, from which the final model error was determined. The quantified mean absolute error and its standard deviation illustrated the model's prediction quality, generalization capabilities, and reproducibility.

For each sample, the model received two spectrograms as input: one for the left channel and one for the right channel. The rationale behind the application of Convolutional Neural Networks (CNNs) to this task was to automatically extract local features from the spectrograms and use these features to predict two contiguous ensemble parameters: ensemble location and width, both measured in degrees. The prediction errors, calculated as the difference between the actual ensemble parameters (known from the binaural synthesis described in Section \ref{subsec:synthesis}) and the predicted values, were utilized in the Adam learning algorithm \parencite{kingma_adam_2014}. During training, the losses for both outputs were combined by summing them, thereby treating both ensemble features with equal importance.

All computations in this study were conducted on a workstation equipped with an RTX Nvidia GeForce 4090 GPU and a 48-core AMD Ryzen ThreadRipper processor. The binaural recordings were synthesized using the MATLAB \parencite{MATLAB} environment with the Audio Toolbox \parencite{MATLAB_Audio_Toolbox}. Feature extraction and model training were performed in the Python \parencite{python}, utilizing the SciPy \parencite{2020SciPy-NMeth} library for feature extraction and Keras for training the CNN model. The complete source code for all stages of the process has been made available by the authors and is accessible to the public at GitHub \parencite{github_repo}. The feature extraction phase required 21 minutes, data partitioning took 34 minutes, and the total training time for all iterations was 40 minutes. In total, the entire training and evaluation process required 95 minutes.

@@

\begin{longtblr}[
  caption = {List of HRTF sets used to synthesize binaural audio excerpts},
  label = {table:hrtfs}
  ]{
  colspec={c|c|>{\raggedright\arraybackslash}X|c|>{\raggedright\arraybackslash}X|c},
  hlines = {},
  vlines = {},
  cell{2}{5} = {r=2}{m},
  cell{2}{6} = {r=2}{m},
  cell{4}{5} = {r=4}{m},
  cell{4}{6} = {r=4}{m},
  cell{8}{5} = {r=3}{m},
  cell{8}{6} = {r=3}{m},
  cell{11}{6} = {r=2}{m},
  cell{13}{5} = {r=3}{m},
  cell{13}{6} = {r=3}{m},
  cell{16}{5} = {r=2}{m},
  cell{16}{6} = {r=2}{m},
  cell{18}{5} = {r=2}{m},
  cell{18}{6} = {r=2}{m},
  cell{20}{5} = {r=3}{m},
  cell{20}{6} = {r=3}{m},
  cell{23}{5} = {r=3}{m},
  cell{23}{6} = {r=3}{m},
  cell{28}{5} = {r=2}{m},
  cell{28}{6} = {r=2}{m},
  cell{30}{5} = {r=2}{m},
  cell{30}{6} = {r=2}{m},

@@

@@

  1.           & Human         & Human subject                             & 1.2                     & RWTH Aachen University \parencite{braren_high-resolution_2020}                                                                             & AACHEN           \\

@@

  3.           & Human         & Subject 2                                 & 1.2                     & Austrian Academy of Sciences \parencite{noauthor_hrtf-database_nodate}                                                                     & ARI              \\

@@

@@

  7.           & Human         & Subject 012                               & 1                       & CIPIC Interface Laboratory, University of California \parencite{algazi_cipic_2001}                                                         & CIPIC            \\

@@

  10.          & Artificial    & Neumann KU 100                            & 0.9                     & NASA (2007) \parencite{andreopoulou_inter-laboratory_2015}                                                                                 & CLUBFRITZ        \\
  11.          & Artificial    & Neumann KU 100                            & 1.5                     & Helsinki University of Technology (2009) \parencite{andreopoulou_inter-laboratory_2015}                                                    &                  \\
  12.          & Artificial    & FABIAN                                    & 1.47                    & Technical University Berlin, Huawei Technologies, Munich Research Centre, Sennheiser Electronic \parencite{brinkmann_cross-evaluated_2019} & HUTUBS           \\

@@

  15.          & Human         & Subject 1003                              & 1.95                    & IRCAM, AKG \parencite{noauthor_listen_2023}                                                                                                & LISTEN           \\

@@

\end{longtblr}
